---
title: "Assignment 3: K Means Clustering"
---

In this assignment we will be applying the K-means clustering algorithm we looked at in class. At the following link you can find a description of K-means:

https://www.cs.uic.edu/~wilkinson/Applets/cluster.html


```{r}
install.packages("ggmap")
library(ggmap)
library(tidyr)
library(dplyr)
library(ggplot2)
library(klaR)
```

Now, upload the file "Class_Motivation.csv" from the Assignment 3 Repository as a data frame called "K1""
```{r}

K1 <- read.csv("Class_Motivation.csv")

```

This file contains the self-reported motivation scores for a class over five weeks. We are going to look for patterns in motivation over this time and sort people into clusters based on those patterns.

But before we do that, we will need to manipulate the data frame into a structure that can be analyzed by our clustering algorithm.

The algorithm will treat each row as a value belonging to a person, so we need to remove the id variable.

```{r}

K2 <- K1[,-1]

```

Q1: It is important to think about the meaning of missing values when clustering. We could treat them as having meaning or we could remove those people who have them. Neither option is ideal. What problems do you foresee if we recode or remove these values? Write your answers below:

A1: If there are a lot of obersvations with missing values relative to total observations, removing them would make the cluster analysis less accurate. K-means clustering is sensitive to oultiers so this is especially true if removal creates artificial ones. Recoding missing values to expected ones reinforces the existing cluster pattern and removes the possibility of uncovering a more accurate one, ultimately compromising the validity of the model. 

We will remove people with missing values for this assignment, but keep in mind the issues that you have identified.


```{r}

K3 <- na.omit(K2) #This command creates a data frame with only those people with no missing values. It "omits" all rows with missing values, also known as a "listwise deletion". EG - It runs down the list deleting rows as it goes.

```

Another pre-processing step used in K-means is to standardize the values so that they have the same range. We do this because we want to treat each week as equally important - if we do not standardise then the week with the largest range will have the greatest impact on which clusters are formed. We standardise the values by using the "scale()" command.

```{r}

K3 <- scale(K3)

```


Now we will run the K-means clustering algorithm we talked about in class. 
1) The algorithm starts by randomly choosing some starting values 
2) Associates all observations near to those values with them
3) Calculates the mean of those clusters of values
4) Selects the observation closest to the mean of the cluster
5) Re-associates all observations closest to this observation
6) Continues this process until the clusters are no longer changing

Notice that in this case we have 5 variables and in class we only had 2. It is impossible to vizualise this process with 5 variables.

Also, we need to choose the number of clusters we think are in the data. We will start with 2.

```{r}

fit <- kmeans(K3, 2)

#We have created an object called "fit" that contains all the details of our clustering including which observations belong to each cluster.

#We can access the list of clusters by typing "fit$cluster", the top row corresponds to the original order the rows were in. Notice we have deleted some rows.



#We can also attach these clusters to the original dataframe by using the "data.frame" command to create a new data frame called K4.

K4 <- data.frame(K3, fit$cluster)

#Have a look at the K4 dataframe. Lets change the names of the variables to make it more convenient with the names() command.

names(K4) <- c("1","2","3","4","5","cluster")
```

Now we need to visualize the clusters we have created. To do so we want to play with the structure of our data. What would be most useful would be if we could visualize average motivation by cluster, by week. To do this we will need to convert our data from wide to long format. Remember your old friends tidyr and dplyr!

First lets use tidyr to convert from wide to long format.
```{r}

K5 <- gather(K4,"week","motivation",1:5)
```

Now lets use dplyr to average our motivation values by week and by cluster.

```{r}

K6 <- K5 %>%
  group_by(week, cluster)
  
K6 <- summarise(K6,avg=mean(motivation))

```

Now it's time to do some visualization: 

https://www.cs.uic.edu/~wilkinson/TheGrammarOfGraphics/GOG.html

And you can see the range of available graphics in ggplot here:

http://ggplot2.tidyverse.org/reference/index.html

We are going to create a line plot similar to the one created in the school dropout paper we looked at in class (Bowers, 2010). It will have motivation on the Y-axis and weeks on the X-axis. To do this we will want our weeks variables to be treated as a number, but because it was created from a variable name it is currently being treated as a character variable. You can see this if you click on the arrow on the left of K6 in the Data pane. Week is designated by "chr". To convert it to numeric, we use the as.numeric command. 

Likewise, since "cluster" is not numeric but rather a categorical label we want to convert it from an "integer" format to a "factor" format so that ggplot does not treat it as a number. We can do this with the as.factor() command.

```{r}

K6$week <- as.numeric(K6$week)

K6$cluster <- as.factor(K6$cluster)

```

Now we can plot our line plot using the ggplot command, "ggplot()".

- The first argument in a ggplot is the dataframe we are using: K6
- Next is what is called an aesthetic (aes), the aesthetic tells ggplot which variables to use and how to use them. Here we are using the variables "week" and "avg" on the x and y axes and we are going color these variables using the "cluster" variable
- Then we are going to tell ggplot which type of plot we want to use by specifiying a "geom()", in this case a line plot: geom_line()
- Finally we are going to clean up our axes labels: xlab("Week") & ylab("Average Motivation")

```{r}
ggplot(K6, 
       aes(week,avg, 
           color = cluster))+ 
  geom_line()+
  theme(plot.title = element_text(hjust = 0.5,
                                  family = "Helvetica",
                                  face="bold"))+
  labs(x ="Week",
       y ="Average Motivation",
       title="Average Motivation per Week")
```

Q2: What patterns do you see in the plot?

A2: The 2 cluster plot (Fig. 1) shows more distinct characteristics (cluster 1=lower motivation; cluster 2=higher motivation) between clusters. The 3 cluster plot (Fig. 2) shows some overlap between groups but indicates motivation for the high motivation group (cluster 1) declines after week 4 while that for the moderate motivation group (cluster 3) increases. Also motivation for clusters 1 & 2 seem to fluctuate together whereas cluster 3 is the opposite up to week 4.

It would be useful to determine how many people are in each cluster. We can do this easily with dplyr.

```{r}
K7 <- count(K4, cluster)
K7
```

Q3: Look at the number of people in each cluster, now repeat this process for 3 rather than 2 clusters. Which cluster grouping do you think is more informative? Write your answer below:

A3: 3 clusters (7, 7, 9); 2 clusters (7, 16)
The number of people per cluster is more evenly distributed in the 3 cluster grouping than 2. Although 2 clusters (Fig.1) provide groups with distinct characteristics, 3 clusters (Fig. 2) could be more informative because it shows distinct changes in motivation after week 4 for highly and moderately motivated students (clusters 1 & 3). This information could be used to address student needs to increase and mantain motivation for the high and low motivation groups (clusters 1 & 2) moving forward.

Once you have done this, save both of your plots to the Assignment 5 file. Create a Zotero item to index your new computer program (Assignment 5.rmd) in Zotero. Then commit your assignment, push it to your Github account and then Pull request your version to the original assignment version so I can see it.

##Part II

Now, try to do the same for the data collected in class. Create two sets of clusters, the answers to the questions and regions where people grew up. 

#Location Clusters

Data Management (Class Activity 6)
```{r}
#Load data
DF1 <- read.csv("HUDK405018-cluster.csv", header = TRUE)
  
#Wrangle data
DF1.1 <- DF1[-1:-2,-1:-17]
DF1.1[DF1.1==""] <- NA
DF1.1 <- DF1.1[complete.cases(DF1.1), ]

#Assign more descriptive column names
colnames(DF1.1) <- c("City", "Country", "Code", "Write", "Acd Read/Vids", "Vids4Fun", "Cook", "SNS", "Socialize")
```

Find lattitudes & longitudes for cities (Class Activity 6)
```{r}
#Request lattitude and longitude from Google Maps API
DF2 <- geocode(as.character(DF1.1$City), output = "latlon", source = "dsk")
```

Form Location Clusters (Class Activity 6)
```{r}
#Create kmeans clusters and attaches them to new data frame DF2.1
locfit <- kmeans(DF2, 4) 
DF2.1 <- data.frame(DF2, locfit$cluster)

#Easier column names
names(DF2.1) <- c("long", "lat", "cluster") 

#Cluster colors are more distinct as factors
DF2.1$cluster <- as.factor(DF2.1$cluster)

#Plot region clusters
ggplot(DF2.1, 
       aes(x = DF2.1$long,
           y = DF2.1$lat,
           color = DF2.1$cluster,
           shape = DF2.1$cluster))+
  geom_point(size = 4)+
  scale_shape_discrete(solid=F)+
  theme(plot.title = element_text(hjust = 0.5,
                                  family = "Helvetica",
                                  face="bold"))+
  labs(x = "Longitude",
       y = "Latitude",
       color = "Region Cluster",
       shape = "Region Cluster",
       title = "Region Clusters by Location")
```

#Activity CLusters

Create Activity Data Frame
```{r}
#Omit City and Country columns
DF3 <- DF1.1[,-1:-2]

#Convert time values to numeric
columns <- names(DF3)
DF3[, columns] <- lapply(columns, function(x) as.numeric(DF3[[x]]))
```

Form Activity Clusters
```{r}
#Create k means clusters by activity
actfit <- kmeans(DF3, 4)

#Attach activity clusters to data frame DF3.1
DF3.1 <- data.frame(DF3, actfit$cluster)

#Assign easer column names
colnames(DF3.1) <- c(c(1:7),"cluster")

#Convert data to long format
DF3.2 <- gather(DF3.1,'act','time',1:7)

#Average times by activity and activity cluster
DF3.3 <- DF3.2 %>% group_by(act,cluster)
DF3.3 <- summarise(DF3.3,avg=mean(time))

#Convert data to desirable class
DF3.3$act <- as.numeric(DF3.3$act)
DF3.3$cluster <- as.factor(DF3.3$cluster)

#Plot activity clusters
ggplot(DF3.3,
       aes(act,avg,color=cluster))+
  geom_line()+
  theme(plot.title = element_text(hjust = 0.5, 
                                  family = "Helvetica", 
                                  face="bold"))+
  labs(x = "Activity:1=Code, 2=Write, 3=Acd Read/Vids, 4=Vids4Fun, 5=Cooking, 6=SNS, 7=Going Out",
       y = "Average Time (hr)",
       title = "Average Activity Time per Cluster",
       color = "Cluster\n")
```

##Part III

Create a visualization that shows the overlap between the two groups of clusters you created in part III.
```{r}
#Combines DF2.1 and DF3.1 into DF4
DF4 <- data.frame(DF2.1,DF3.1)

#Change column names
colnames(DF4)[3] <-"loc_cluster"
colnames(DF4)[11] <-"act_cluster"

#Convert to factor
DF4$act_cluster <- as.factor(DF4$act_cluster)
DF4$loc_cluster <- as.factor(DF4$loc_cluster)

#Plot Region and Activity Clusters by Location
ggplot(DF4,
       aes(x = long,
           y = lat,
           color = DF4$loc_cluster,
           shape = DF4$act_cluster))+
  geom_point(size = 4)+
  scale_shape_discrete(solid=F)+
  theme(plot.title = element_text(hjust = 0.5,
                                  family = "Helvetica",
                                  face="bold"))+
  labs(x = "Longitude",
       y = "Latitude",
       title = "Region and Activity Clusters by Location",
       color = "Region Clusters",
       shape = "Activity Clusters")
```